Motivating Exploration in Reinforcement Learning

### 2020-04-16 paper/main.tex
discussion

\subsubsubsection{\textbf{Pong}}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/pong}
  \caption{With Exploration}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/pong_no_exp}
  \caption{Without Exploration}
\end{subfigure}
\caption{Pong environment visualizations}
\label{fig:pong}
\end{figure}
With regards to Pong, the environment with the sparsest rewards from the three that are investigated in this paper, both agents without and with exploration respectively still make use of movement of the paddle to maximize cumulative reward as can be seen in Fig \ref{fig:pong}. However, the agent with exploration noticeably has a much higher rate of movement compared to the agent without exploration, which could lead to more states of the environment being explored. The agent with exploration often makes many movements that are much faster than the agent without exploration, as can be observed in the supplementary videos.

\subsubsubsection{\textbf{Breakout}}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/breakout}
  \caption{With Exploration}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/breakout_no_exp}
  \caption{Without Exploration}
\end{subfigure}
\caption{Breakout environment visualizations}
\label{fig:breakout}
\end{figure}
With regards to Breakout, the environment out of the three investigated in this paper with a higher reward density than Pong but a lower reward density than Seaquest, there is a very noticeable difference in the behaviour of the two agents. The agent without exploration tends to move to the rightmost corner, and simply stay there for most of the game’s duration as can be seen in Fig. \ref{fig:breakout}. The agent with exploration makes movements all across the screen in order to maximize cumulative reward, and intuitively, explores its environment much more than the agent without exploration. This is a very distinct difference in behaviour that can be clearly observed, especially in the supplementary videos provided.

\subsubsubsection{\textbf{Seaquest}}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/seaquest}
  \caption{With Exploration}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/seaquest_no_exp}
  \caption{Without Exploration}
\end{subfigure}
\caption{Seaquest environment visualizations}
\label{fig:seaquest}
\end{figure}
With regards to Seaquest, the environment with the highest reward density, there is also a very noticeable difference in the behaviour of the agents as seen in Fig. \ref{fig:seaquest}. This difference between the two agents is the same as the difference noticed in the agents of Breakout. The agent without exploration in Seaquest does not explore its environment at all, and instead opts to stay above sea level at the top left corner. This behaviour is quite unique because it completely avoids any negative rewards in the environment caused by the fish attacking the agent or due to oxygen loss. On the other hand, the agent with exploration makes movements to fully explore the environment to maximize reward while taking the risk of negative rewards as well.

The differences in behaviour noticed in the agents with and without exploration of both Breakout and Seaquest could perhaps be explained by the lack of maximizing the entropy of the next-action sampling distribution in the agent without exploration. Without this maximization of entropy of the next-action sampling distribution, there is no degree of randomness in the policy network’s state to action mapping. In turn, the lack of randomness may be causing a collapse in the explorative behaviour due to some, perhaps static, actions having a very high probability defined by the policy network logits.

@@

In Pong, it is observed that the agent waits for the ball for a bit, and then rapidly moves towards the ball to hit it with the edge of the paddle in order to launch the ball quickly off of the paddle with a high velocity. Perhaps this is a strategy used by the agent to cause a rapid change in velocity that causes the opponent to be unable to catch it in time due to slower reaction speeds.

In Breakout, it is observed that the agent tends to hit the ball from the rightmost corner if it can. This might be a strategy used to break apart the bricks at the top left corner in order to push the ball on top of the bricks, and consequently destroy many more bricks from above in order to accumulate much higher rewards.

In Seaquest, it is observed that the agent takes into consideration the oxygen bar, which is a purely visual indicator. The agent decides the next actions based on their oxygen remaining as well, opting to return to the surface of the water to regain oxygen in order to further extend the opportunity to accumulate more rewards.

\section{Discussion}

>>>>

\subsubsubsection{\textbf{Pong}}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/pong}
  \caption{With Exploration}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/pong_no_exp}
  \caption{Without Exploration}
\end{subfigure}
\caption{Pong environment visualizations}
\label{fig:pong}
\end{figure}
With regards to \texttt{Pong}, the environment with the sparsest rewards from the three that are investigated in this paper, both agents without and with exploration respectively still make use of movement of the paddle to maximize cumulative reward as can be seen in Fig \ref{fig:pong}. However, the agent with exploration noticeably has a much higher rate of movement compared to the agent without exploration, which could lead to more states of the environment being explored. The agent with exploration often makes many movements that are much faster than the agent without exploration, as can be observed in the supplementary videos.

\subsubsubsection{\textbf{Breakout}}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/breakout}
  \caption{With Exploration}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/breakout_no_exp}
  \caption{Without Exploration}
\end{subfigure}
\caption{Breakout environment visualizations}
\label{fig:breakout}
\end{figure}
With regards to \texttt{Breakout}, the environment out of the three investigated in this paper with a higher reward density than \texttt{Pong} but a lower reward density than \texttt{Seaquest}, there is a very noticeable difference in the behaviour of the two agents. The agent without exploration tends to move to the rightmost corner, and simply stay there for most of the game’s duration as can be seen in Fig. \ref{fig:breakout}. The agent with exploration makes movements all across the screen in order to maximize cumulative reward, and intuitively, explores its environment much more than the agent without exploration. This is a very distinct difference in behaviour that can be clearly observed, especially in the supplementary videos provided.

\subsubsubsection{\textbf{Seaquest}}
\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/seaquest}
  \caption{With Exploration}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{figures/seaquest_no_exp}
  \caption{Without Exploration}
\end{subfigure}
\caption{Seaquest environment visualizations}
\label{fig:seaquest}
\end{figure}
With regards to \texttt{Seaquest}, the environment with the highest reward density, there is also a very noticeable difference in the behaviour of the agents as seen in Fig. \ref{fig:seaquest}. This difference between the two agents is the same as the difference noticed in the agents of \texttt{Breakout}. The agent without exploration in Seaquest does not explore its environment at all, and instead opts to stay above sea level at the top left corner. This behaviour is quite unique because it completely avoids any negative rewards in the environment caused by the fish attacking the agent or due to oxygen loss. On the other hand, the agent with exploration makes movements to fully explore the environment to maximize reward while taking the risk of negative rewards as well.

The differences in behaviour noticed in the agents with and without exploration of both \texttt{Breakout} and \texttt{Seaquest} could perhaps be explained by the lack of maximizing the entropy of the next-action sampling distribution in the agent without exploration. Without this maximization of entropy of the next-action sampling distribution, there is no degree of randomness in the policy network’s state to action mapping. In turn, the lack of randomness may be causing a collapse in the explorative behaviour due to some, perhaps static, actions having a very high probability defined by the policy network logits.

@@

In \texttt{Pong}, it is observed that the agent waits for the ball for a bit, and then rapidly moves towards the ball to hit it with the edge of the paddle in order to launch the ball quickly off of the paddle with a high velocity. Perhaps this is a strategy used by the agent to cause a rapid change in velocity that causes the opponent to be unable to catch it in time due to slower reaction speeds.

In \texttt{Breakout}, it is observed that the agent tends to hit the ball from the rightmost corner if it can. This might be a strategy used to break apart the bricks at the top left corner in order to push the ball on top of the bricks, and consequently destroy many more bricks from above in order to accumulate much higher rewards.

In \texttt{Seaquest}, it is observed that the agent takes into consideration the oxygen bar, which is a purely visual indicator. The agent decides the next actions based on their oxygen remaining as well, opting to return to the surface of the water to regain oxygen in order to further extend the opportunity to accumulate more rewards.

\section{Discussion}
Even with the massive computational resources available today, artificial agents face difficulty in balancing explicit exploration of the environment and stable convergence towards a useful, learned policy. In this work, we explored a general formulation of curiosity that aims to motivate the exploration in reinforcement learning agents while remaining invariant to the algorithm and problem in question. This general formulation of curiosity was applied to the A2C RL algorithm in various environments with different extrinsic reward densities to examine the effects of an entropy term (of the policy action probability distribution) and a curiosity module providing an intrinsic reward signal.

As discussed in the results section above, the curiosity-driven (ICM) agents in \texttt{Pong} learn and train much faster than the entropy-based agents, which was explained by the difference in the style of exploration. Moreover, the baseline A2C model with no exploration performed poorly due to the fact that the agent was not exploring at all. With regards to \texttt{Breakout} and \texttt{Seaquest}, particularly dense reward environments, the ICM performance remains an area of improvement through hyper-parameter tuning. The baseline A2C models perform poorly in these environments as discussed previously, but they perform better in these environments relative to sparse reward environments like \texttt{Pong}. Our results show that exploration factors applied to baseline algorithms can improve model performance and decrease training time through intrinsic reward signals. In particular, exploration is crucial across sparse reward environments, which we wanted to verify and which is backed by our results.

The limitations we faced in our work were mostly related to time available to complete this work. With more time, we would have liked to train all of  our agents for longer periods of time until convergence. Moreover, we would prefer to have done more experiments with more time to observe the quantitative effects of hyperparameter tuning for specific environments such as \texttt{Breakout} and \texttt{Seaquest} and agents with intrinsic curiosity (ICM) since we were unable to do so in the time we had.

In the future, we would like to diversify the environments investigated in this paper. \texttt{Pong}, \texttt{Breakout}, and \texttt{Seaquest} are all Atari games where the states and frames of the game do not change drastically from one state to another. This could perhaps be a factor affecting the ICM due to the fact that novelty in states for curiosity is defined by the change in its features. Through more experiments in a diversified set of environments including environments with more distinct states, additional insights about the effect of exploration and curiosity might be revealed.
